A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-23 paper.Rmd
working hard on revising the experimental setup section

The power of a conventional hypothesis test is affected by both the true parameter $\boldsymbol{\theta}$ and the sample size $n$. These two can be quantified in terms of effect size $E$ to measure the strength of the residual departures from the model assumptions. Details about the effect size are provided in Section \ref{effect-size} after the introduction of the simulation model used in our experiment. The theoretical power of a test is sometimes not a trival solution, but it can be estimated if the data generating process is known. We use a predefined model to generate a large set of simulated data under different effect sizes, and record if the conventional test rejects $H_0$. The probability of the conventional test rejects $H_0$ is then fitted by a logistic regression formulated as

@@

### Controlling the signal strength of the signal

@@

Effect size in statistics measures the strength of the signal relative to the noise. It is surprisingly difficult to quantify in general, even for simulated data as used in this experiment.

@@

$$E = \frac{1}{2}\left(\boldsymbol{\mu}_z'(diag(\boldsymbol{R}\sigma^2))^{-1}\boldsymbol{\mu}_z\right)$$
\noindent where $diag(.)$ is the diagonal matrix constructed from the diagonal elements of a matrix,
$\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{H}$ is the residual operator, $\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'$ is the hat matrix, $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$ is the expected values of residuals with $\boldsymbol{Z}$ be any higher order terms of $\boldsymbol{X}$ leave out by the regression equation and $\boldsymbol{\beta}_z$ be the corresponding coefficients, and $\sigma^2\boldsymbol{I}$ is the assumed covariance matrix of the error term when $H_0$ is true.

@@

$$E = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')|}{|diag(\boldsymbol{R})|} - n + tr(diag(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')^{-1}diag(\boldsymbol{R}))\right)$$
\noindent where $\boldsymbol{V}$ is the actual covariance matrix of the error term. (Derivations for these equations are provided in the Appendix.)

@@

As shown in Table \ref{tab:model-factor-table}, there are a total of $4 \times 4 \times 3 \times 4 = 192$ and $3 \times 5 \times 3 \times 4 = 180$ number of combinations of parameter values for non-linearity model and heteroskedasticity model respectively. Three replications are made for each of the combination results in $192 \times 3 = 576$ and $180 \times 3 = 540$ lineups. In addition, each lineup is designed to be evaluated by five different participants. After attempting some pilot studies internally, we decide to present a block of 20 lineups to every participant. And to ensure the quality of the survey data, two lineups with obvious visual patterns are included as attention checks. Thus, $576 \times 5 / (20-2) = 160$ and $540 \times 5 / (20-2) = 150$ participants are recruited to satisfy the design of the data collection period I and II respectively.

>>>>

The power of a conventional hypothesis test is affected by both the true parameter $\boldsymbol{\theta}$ and the sample size $n$. These two can be quantified in terms of effect size $E$ to measure the strength of the residual departures from the model assumptions. Details about the calculation of effect size are provided in Section \ref{effect-size} after the introduction of the simulation model used in our experiment. The theoretical power of a test is sometimes not a trival solution, but it can be estimated if the data generating process is known. We use a predefined model to generate a large set of simulated data under different effect sizes, and record if the conventional test rejects $H_0$. The probability of the conventional test rejects $H_0$ is then fitted by a logistic regression formulated as

@@

### Controlling the signal strength

@@

Effect size in statistics measures the strength of the signal relative to the noise. It is surprisingly difficult to quantify, even for simulated data as used in this experiment.

@@

\begin{equation}
E = \frac{1}{2}\left(\boldsymbol{\mu}_z'(diag(\boldsymbol{R}\sigma^2))^{-1}\boldsymbol{\mu}_z\right)
\label{eq:nonlin-effect-size}
\end{equation}

\noindent where $diag(.)$ is the diagonal matrix constructed from the diagonal elements of a matrix,
$\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{H}$ is the residual operator, $\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'$ is the hat matrix, $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$ is the expected values of residuals where $\boldsymbol{Z}$ contain any higher order terms of $\boldsymbol{X}$ left out of the regression equation, $\boldsymbol{\beta}_z$ contains the corresponding coefficients, and $\sigma^2\boldsymbol{I}$ is the assumed covariance matrix of the error term when $H_0$ is true.

@@

\begin{equation}
E = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')|}{|diag(\boldsymbol{R})|} - n + tr(diag(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')^{-1}diag(\boldsymbol{R}))\right)
\label{eq:hetero-effect-size}
\end{equation}

\noindent where $\boldsymbol{V}$ is the actual covariance matrix of the error term. Derivations for these equations are provided in the Appendix.

@@

From Table \ref{tab:model-factor-table} it can be computed that there are a total of $4 \times 4 \times 3 \times 4 = 192$ and $3 \times 5 \times 3 \times 4 = 180$ combinations of parameter values for the non-linearity model and heteroskedasticity models respectively. Three replications for each combination results in $192 \times 3 = 576$ and $180 \times 3 = 540$ lineups, respectively.

Each lineup needs to be evaluated by at least five participants. From previous work, and additional pilot studies for this experiment, we decided to record evaluations from 20 lineups for each participant. Two of the 20 lineups with clear visual patterns were used as attention checks, to ensure quality data. Thus, $576 \times 5 / (20-2) = 160$ and $540 \times 5 / (20-2) = 150$ participants were needed to cover the experimental design for the data collection periods I and II, respectively. The factor levels and range of difficulty was assigned relatively equally among participants. The Appendix contains graphical summaries of these assignments for each subject.

### 2023-04-23 paper.tex
working hard on revising the experimental setup section

The power of a conventional hypothesis test is affected by both the true
parameter \(\boldsymbol{\theta}\) and the sample size \(n\). These two
can be quantified in terms of effect size \(E\) to measure the strength
of the residual departures from the model assumptions. Details about the
effect size are provided in Section \ref{effect-size} after the
introduction of the simulation model used in our experiment. The
theoretical power of a test is sometimes not a trival solution, but it
can be estimated if the data generating process is known. We use a
predefined model to generate a large set of simulated data under
different effect sizes, and record if the conventional test rejects
\(H_0\). The probability of the conventional test rejects \(H_0\) is
then fitted by a logistic regression formulated as

@@

\hypertarget{controlling-the-signal-strength-of-the-signal}{%
\subsubsection{Controlling the signal strength of the
signal}\label{controlling-the-signal-strength-of-the-signal}}

@@

\[E = \frac{1}{2}\left(\boldsymbol{\mu}_z'(diag(\boldsymbol{R}\sigma^2))^{-1}\boldsymbol{\mu}_z\right)\]
\noindent where \(diag(.)\) is the diagonal matrix constructed from the
diagonal elements of a matrix,
\(\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{H}\) is the residual
operator,
\(\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the hat matrix,
\(\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z\)
is the expected values of residuals with \(\boldsymbol{Z}\) be any
higher order terms of \(\boldsymbol{X}\) leave out by the regression
equation and \(\boldsymbol{\beta}_z\) be the corresponding coefficients,
and \(\sigma^2\boldsymbol{I}\) is the assumed covariance matrix of the
error term when \(H_0\) is true.

@@

\[E = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')|}{|diag(\boldsymbol{R})|} - n + tr(diag(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')^{-1}diag(\boldsymbol{R}))\right)\]
\noindent where \(\boldsymbol{V}\) is the actual covariance matrix of
the error term. (Derivations for these equations are provided in the
Appendix.)

@@

As shown in Table \ref{tab:model-factor-table}, there are a total of
\(4 \times 4 \times 3 \times 4 = 192\) and
\(3 \times 5 \times 3 \times 4 = 180\) number of combinations of
parameter values for non-linearity model and heteroskedasticity model
respectively. Three replications are made for each of the combination
results in \(192 \times 3 = 576\) and \(180 \times 3 = 540\) lineups. In
addition, each lineup is designed to be evaluated by five different
participants. After attempting some pilot studies internally, we decide
to present a block of 20 lineups to every participant. And to ensure the
quality of the survey data, two lineups with obvious visual patterns are
included as attention checks. Thus, \(576 \times 5 / (20-2) = 160\) and
\(540 \times 5 / (20-2) = 150\) participants are recruited to satisfy
the design of the data collection period I and II respectively.

>>>>

The power of a conventional hypothesis test is affected by both the true
parameter \(\boldsymbol{\theta}\) and the sample size \(n\). These two
can be quantified in terms of effect size \(E\) to measure the strength
of the residual departures from the model assumptions. Details about the
calculation of effect size are provided in Section \ref{effect-size}
after the introduction of the simulation model used in our experiment.
The theoretical power of a test is sometimes not a trival solution, but
it can be estimated if the data generating process is known. We use a
predefined model to generate a large set of simulated data under
different effect sizes, and record if the conventional test rejects
\(H_0\). The probability of the conventional test rejects \(H_0\) is
then fitted by a logistic regression formulated as

@@

\hypertarget{controlling-the-signal-strength}{%
\subsubsection{Controlling the signal
strength}\label{controlling-the-signal-strength}}

@@

\begin{equation}
E = \frac{1}{2}\left(\boldsymbol{\mu}_z'(diag(\boldsymbol{R}\sigma^2))^{-1}\boldsymbol{\mu}_z\right)
\label{eq:nonlin-effect-size}
\end{equation}

\noindent where \(diag(.)\) is the diagonal matrix constructed from the
diagonal elements of a matrix,
\(\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{H}\) is the residual
operator,
\(\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the hat matrix,
\(\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z\)
is the expected values of residuals where \(\boldsymbol{Z}\) contain any
higher order terms of \(\boldsymbol{X}\) left out of the regression
equation, \(\boldsymbol{\beta}_z\) contains the corresponding
coefficients, and \(\sigma^2\boldsymbol{I}\) is the assumed covariance
matrix of the error term when \(H_0\) is true.

@@

\begin{equation}
E = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')|}{|diag(\boldsymbol{R})|} - n + tr(diag(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')^{-1}diag(\boldsymbol{R}))\right)
\label{eq:hetero-effect-size}
\end{equation}

\noindent where \(\boldsymbol{V}\) is the actual covariance matrix of
the error term. Derivations for these equations are provided in the
Appendix.

@@

From Table \ref{tab:model-factor-table} it can be computed that there
are a total of \(4 \times 4 \times 3 \times 4 = 192\) and
\(3 \times 5 \times 3 \times 4 = 180\) combinations of parameter values
for the non-linearity model and heteroskedasticity models respectively.
Three replications for each combination results in
\(192 \times 3 = 576\) and \(180 \times 3 = 540\) lineups, respectively.

Each lineup needs to be evaluated by at least five participants. From
previous work, and additional pilot studies for this experiment, we
decided to record evaluations from 20 lineups for each participant. Two
of the 20 lineups with clear visual patterns were used as attention
checks, to ensure quality data. Thus, \(576 \times 5 / (20-2) = 160\)
and \(540 \times 5 / (20-2) = 150\) participants were needed to cover
the experimental design for the data collection periods I and II,
respectively. The factor levels and range of difficulty was assigned
relatively equally among participants. The Appendix contains graphical
summaries of these assignments for each subject.
